---
title: "Genome_summary"
output: html_document
date: "2026-02-23"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(width=200)

```

# Genome summary

This document contains the code for running  the statistical analyses and data visualization for the nodule trapping data.

# Load required libraries

```{r Load libraries, results='hide', message=FALSE, warning=FALSE}

library(lme4)
library(phyloseq)
library(car)
library(emmeans)
library(ggplot2)
library(patchwork)
library(vegan)
library(dplyr)
library(tidyr)
library(kableExtra)
library(tibble)
library(tidyverse)
library(UpSetR)
library(gplots)
library(gggenes)
library(RColorBrewer)

```

# Statistical analysis of the genome metrics

Examine if genome size and replicon number differs significantly across specie

```{r Statistical analysis - load data, results='hide', message=FALSE, warning=FALSE}

# Load the data
data <- read.table("genome_stats.txt",
                   header = TRUE, row.names = NULL)

```

First check if genome size is statistically different across species

```{r Statistical analysis - genome size, message=FALSE, warning=FALSE}

# Perform statistical analysis on genome size
model.size <- lm(genome_size ~ Species, data=data)
summary(model.size)
#model.size.aov <- Anova(model.size, type = 2)
#model.size.em <- emmeans(model.size, pairwise ~ Species,
#                         infer = TRUE, type = "response")
#model.size.em$emmeans
#model.size.em$contrasts

```

Genome size is statistically different across species.

Now check if the number of replicons differs, limited to only complete genomes

```{r Statistical analysis - contig number, rmessage=FALSE, warning=FALSE}

# Perform statistical analysis on replicon count
data.complete <- data[data$genome_status == "complete", ]
model.replicon <- lm(contig_count ~ Species, data=data.complete)
summary(model.replicon)
#model.replicon.aov <- Anova(model.replicon, type = 2)
#model.replicon.em <- emmeans(model.replicon, pairwise ~ Species,
#                         infer = TRUE, type = "response")
#model.replicon.em$emmeans
#model.replicon.em$contrasts

```

Replicon count is statistically different across species.

Now clear the workspace in preparation for next analyses.

```{r Statistical analysis - clear, results='hide', message=FALSE, warning=FALSE}

# Clean the workspace
rm(list = ls())

```

# Summarize genome metrics

Prepare tables with summary statistics for the genome metrics of interest

```{r Genome metrics - load, results='hide', message=FALSE, warning=FALSE}

# Load the data
data <- read.table("genome_stats.txt",
                   header = TRUE, row.names = NULL)

```

Prepare summary statistics for genome size.

```{r Genome metrics - size, message=FALSE, warning=FALSE}

# Get summary metrics for genome size
genome.size.stats <- data %>%
  group_by(Species) %>%
  summarize(Count = n(),
            Mean = round(mean(genome_size, na.rm = TRUE), 0),
            Median = round(median(genome_size, na.rm = TRUE), 0),
            SD = round(sd(genome_size, na.rm = TRUE), 0),
            Min = round(min(genome_size, na.rm = TRUE), 0),
            Max = round(max(genome_size, na.rm = TRUE), 0))

# Create a table rounded to 0 decimal place
kable(genome.size.stats, align = "l",
      caption = "Genome size statistics") %>%
  kable_classic(full_width = FALSE)

```

Now prepare summary statistics for contig number, limiting the analysis to complete genomes

```{r Genome metrics - contigs, message=FALSE, warning=FALSE}

# Get summary metrics for contig count
data.complete <- data[data$genome_status == "complete", ]
replicon.stats <- data.complete %>%
  group_by(Species) %>%
  summarize(Count = n(),
            Mean = round(mean(contig_count, na.rm = TRUE), 0),
            Median = round(median(contig_count, na.rm = TRUE), 0),
            SD = round(sd(contig_count, na.rm = TRUE), 0),
            Min = round(min(contig_count, na.rm = TRUE), 0),
            Max = round(max(contig_count, na.rm = TRUE), 0))

# Create a table rounded to 0 decimal place
kable(replicon.stats, align = "l",
      caption = "Contig count statistics") %>%
  kable_classic(full_width = FALSE)

```

Now, prepare a graph summarizing the genome size distribution

```{r Genome metrics - figure create, message=FALSE, warning=FALSE}

# Prepare a box plot with overlapping datapoints
plot_boxplot <- ggplot(data, aes(x = Species, y = genome_size)) +
  geom_boxplot() +
  geom_jitter(colour="grey", size=1, alpha=1, width=0.25) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Genome size (Mb)",
                     limits = c(6000000,8000000),
                     breaks = c(6000000,6500000,7000000,7500000,8000000),
                     labels = function(x) x / 1e6) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 12, angle = 30, hjust = 1, vjust=1, face="italic"),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12), 
        legend.position = "none", 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        plot.margin = unit(c(0.5,0.5,0.5,2), "cm"))  
plot_boxplot

```

```{r Genome metrics - figure save, results='hide', message=FALSE, warning=FALSE}

# Save as a SVG file
svg(filename = "genome.size.svg", width=8.5, height=5)
plot_boxplot
dev.off()

# Save as a PDF file
pdf("genome.size.pdf", width=8.5, height=5)
plot_boxplot
dev.off()

```

Finally, clear the workspace in preparation for next analyses.

```{r Genome metrics - clear, results='hide', message=FALSE, warning=FALSE}

# Clean the workspace
rm(list = ls())

```

# Average nucleotide identity (ANI)

Make a heatmap clustering the 216 Rhizobium isolates based on pairwise ANI values.

```{r ANI - create, message=FALSE, warning=FALSE}

# Load the data
data <- read.table("ANI_matrix.txt",
                   header = TRUE, row.names = 1)
data <- data.matrix(data)

# Set clustering variables
dist.pear <- function(x) as.dist(1-cor(t(x)))
hclust.ave <- function(x) hclust(x, method="average")

# Prepare heatmap
my_palette <- colorRampPalette(c("white", "light blue", "dark blue"))(n=4)
heatmap.2(data, trace=c("none"), density.info=c("none"), col=my_palette,
          breaks = c(94.9,95,98.9,99.8, 100),
          hclustfun = hclust.ave,
          distfun = dist.pear,
          lmat=matrix(c(4,2,3,1), nrow=2, ncol=2), key=T,
          cexRow=0.1, cexCol=0.1)

```

Now save the figure.

```{r ANI - export, results='hide', message=FALSE, warning=FALSE}

# Export as a SVG file
svg(filename = 'ANI_matrix.svg')
heatmap.2(data, trace=c("none"), density.info=c("none"), col=my_palette,
          breaks = c(94.9,95,98.9,99.8, 100),
          hclustfun = hclust.ave,
          distfun = dist.pear,
          lmat=matrix(c(4,2,3,1), nrow=2, ncol=2), key=T,
          cexRow=0.1, cexCol=0.1)
dev.off()

# Explort as a PDF file
pdf('ANI_matrix.pdf')
heatmap.2(data, trace=c("none"), density.info=c("none"), col=my_palette,
          breaks = c(94.9,95,98.9,99.8, 100),
          hclustfun = hclust.ave,
          distfun = dist.pear,
          lmat=matrix(c(4,2,3,1), nrow=2, ncol=2), key=T,
          cexRow=0.1, cexCol=0.1)
dev.off()

```

Now clear the workspace in preparation for next analyses.

```{r ANI - clear, results='hide', message=FALSE, warning=FALSE}

# Clean the workspace
rm(list = ls())

```

# Pangenome visualization

Create UpSetR plots to visualize the results of the pangenome analysis. First, create species level data where a gene is considered present in a species as long as it shows up at least one, and then create the plot.

```{r Pangenome total - create, message=FALSE, warning=FALSE}

# Load the data 
gene_bean <- read.table("gene_presence_absence.tsv",
                        header = TRUE)

# Clean up the species names in the columns 
gene_bean2 <- gene_bean %>%
  pivot_longer(
    cols = -Gene,             
    names_to = "full_name",
    values_to = "PA",
    values_transform = list(PA = as.numeric)  
  ) %>%
  mutate(Species = str_extract(full_name, "^[^_]+_[^_]+")) %>%
  group_by(Gene, Species) %>%
  summarise(PA = max(PA), .groups = "drop") %>%
  pivot_wider(names_from = Species, values_from = PA)

# Prep data for the upsetr plot 
gene_bean3 <- gene_bean2 %>%
  dplyr::select(-Gene)
  
gene_bean3 <- as.data.frame(gene_bean3)
rownames(gene_bean3) <- NULL

# Prep the names for plot 
colnames(gene_bean3) <- gsub("_", " ", colnames(gene_bean3))

plot1<-upset(
  gene_bean3,
  point.size = 3.5, 
  order.by = "freq",
  sets = colnames(gene_bean3),   
  mainbar.y.label = "Shared gene groups",
  sets.x.label = "Total genes",
  mb.ratio = c(0.6,0.4),
  text.scale = c(1.6, # intersection title
                           1.6, # size tick labels
                           1.6, # set size title
                           1.6, # set tick sizes 
                           1.6, # species labels
                           0.8),
)
plot1

```

```{r Pangenome total - export, results='hide', message=FALSE, warning=FALSE}

# Save the plot 
pdf("UpSetR.total.pdf", width=10, height=6)
plot1
dev.off()

# Clean the workspace
rm(list = ls())

```

Now, create species level data where a gene is considered present in a species only if it is present in at least 95% of the isolates. Then, create a plot limited to genes that are present in the core or soft core genome of at least one species but absent in all other species.

```{r Pangenome core - create, message=FALSE, warning=FALSE}

# Load the data 
gene_bean <- read.table("gene_presence_absence.tsv",
                        header = TRUE)

# Clean up the species names in the columns convert anything in the soft core genome to being in the core genome
gene_bean2 <- gene_bean %>%
  pivot_longer(
    cols = -Gene,             
    names_to = "full_name",
    values_to = "PA",
    values_transform = list(PA = as.numeric)  
  ) %>%
  mutate(Species = str_extract(full_name, "^[^_]+_[^_]+")) %>%
  group_by(Gene, Species) %>%
  summarise(PA = sum(PA) / n(), .groups = "drop") %>%
  #mutate(PA = replace(PA, PA <= 0.05, 0)) %>%
  mutate(PA = replace(PA, PA >= 0.95, 1)) %>%
  filter(PA == 0 | PA == 1) %>%
  pivot_wider(names_from = Species, values_from = PA) %>%
  drop_na()

# Prep data for the upsetr plot 
gene_bean3 <- gene_bean2 %>%
  dplyr::select(-Gene)
  
gene_bean3 <- as.data.frame(gene_bean3)
rownames(gene_bean3) <- NULL

# Prep the names for plot 
colnames(gene_bean3) <- gsub("_", " ", colnames(gene_bean3))

plot1<-upset(
  gene_bean3,
  point.size = 3.5, 
  order.by = "freq",
  sets = colnames(gene_bean3),   
  mainbar.y.label = "Shared gene groups",
  sets.x.label = "Total genes",
  mb.ratio = c(0.6,0.4),
  text.scale = c(1.6, # intersection title
                           1.6, # size tick labels
                           1.6, # set size title
                           1.6, # set tick sizes 
                           1.6, # species labels
                           0.8),
)
plot1

```

```{r Pangenome core - export, results='hide', message=FALSE, warning=FALSE}

# Save the plot 
pdf("UpSetR.core.pdf", width=10, height=6)
plot1
dev.off()

# Clean the workspace
rm(list = ls())

```

# Nod locus

Create an image showing the organization of nod genes in the representative isolates that have complete genomes. Also includ Rhizobium coratiense QUR0161 to have a second representative of the sv. phaseoli gamma-b allele.

```{r Nod locus - complete, message=FALSE, warning=FALSE}

# Load the data
genes <- read.table("nod_locus.txt",
                    header = TRUE, row.names = NULL, sep="\t")
gaps <- read.table("nod_breaks.txt",
                   header = TRUE, row.names = NULL, sep="\t")

# Order the isolates
genes$molecule <- factor(genes$molecule, levels = c("Rhizobium_mongolense_QUR0164","Rhizobium_leguminosarum_QUR0062","Rhizobium_croatiense_QUR0161","Rhizobium_sp_QUR0225","Rhizobium_sp_QUR0188","Rhizobium_sp_QUR0108","Rhizobium_sp_QUR0071","Rhizobium_leguminosarum_QUR0136","Rhizobium_leguminosarum_QUR0050","Rhizobium_croatiense_QUR0171","Rhizobium_croatiense_QUR0087","Rhizobium_leguminosarum_QUR0210","Rhizobium_leguminosarum_QUR0186","Rhizobium_leguminosarum_QUR0177","Rhizobium_laguerreae_QUR0112","Rhizobium_johnstonii_QUR0098","Rhizobium_hidalgonense_QUR0172","Rhizobium_croatiense_QUR0168","Rhizobium_croatiense_QUR0156","Rhizobium_croatiense_QUR0154","Rhizobium_anhuiense_QUR0074","Rhizobium_sophoriradicis_QUR0003","Rhizobium_croatiense_QUR0011"))
gaps$molecule <- factor(gaps$molecule, levels = c("Rhizobium_mongolense_QUR0164","Rhizobium_leguminosarum_QUR0062","Rhizobium_croatiense_QUR0161","Rhizobium_sp_QUR0225","Rhizobium_sp_QUR0188","Rhizobium_sp_QUR0108","Rhizobium_sp_QUR0071","Rhizobium_leguminosarum_QUR0136","Rhizobium_leguminosarum_QUR0050","Rhizobium_croatiense_QUR0171","Rhizobium_croatiense_QUR0087","Rhizobium_leguminosarum_QUR0210","Rhizobium_leguminosarum_QUR0186","Rhizobium_leguminosarum_QUR0177","Rhizobium_laguerreae_QUR0112","Rhizobium_johnstonii_QUR0098","Rhizobium_hidalgonense_QUR0172","Rhizobium_croatiense_QUR0168","Rhizobium_croatiense_QUR0156","Rhizobium_croatiense_QUR0154","Rhizobium_anhuiense_QUR0074","Rhizobium_sophoriradicis_QUR0003","Rhizobium_croatiense_QUR0011"))

# Order the gene colours
genes$gene <- factor(genes$gene, levels = c("nolL", "nodD", "nodA", "nodZ", "nodB", "nodC", "nodS", "nodU", "nodI", "nodJ", "nolE", "noeL", "nolO", "hsnT", "nodF", "nodE", "nodM", "nodH", "nodP", "nodQ"))

# Make the plot 
myColors <- colorRampPalette(brewer.pal(12, "Paired"))(20)
plot1 <- ggplot(genes, aes(xmin = start, xmax = end, 
                  y = molecule, fill = gene, label = gene,
                  forward = orientation)) +
  geom_gene_arrow(arrowhead_height = unit(5, "mm"), arrowhead_width = unit(1, "mm"),
                  arrow_body_height = unit(5, "mm")) +
  geom_gene_label() +
  geom_feature(
    data = gaps,
    aes(x = position, y = molecule)
  ) +
  geom_feature_label(
    data = gaps,
    aes(x = position, y = molecule, label = name)
  ) +
  #facet_wrap(~ molecule, scales = "free", ncol = 1) +
  scale_fill_manual(values = myColors) +
  theme_genes() +
  scale_x_continuous(limits = c(0,26000), breaks = c(0,6500,13000,19500,26000))
plot1

```

```{r Nod locus - complete export, results='hide', message=FALSE, warning=FALSE}

# Save the plot 
pdf("nod.locus.full.pdf", width=10, height=10)
plot1
dev.off()

```

Since many of them look identical, there is no need to have all of them in the main figure. Therefore, create a reduced version with just one representative per type.

```{r Nod locus - subset, message=FALSE, warning=FALSE}

# Filter to a smaller subset so that the image is more manageable
isolates.subset <- c("Rhizobium_mongolense_QUR0164","Rhizobium_croatiense_QUR0161","Rhizobium_croatiense_QUR0087","Rhizobium_croatiense_QUR0168","Rhizobium_croatiense_QUR0011")
filtered.genes <- genes %>% 
  filter(molecule %in% isolates.subset)
filtered.gaps <- gaps %>% 
  filter(molecule %in% isolates.subset)

# Make the plot 
myColors <- colorRampPalette(brewer.pal(12, "Paired"))(20)
plot2 <- ggplot(filtered.genes, aes(xmin = start, xmax = end, 
                  y = molecule, fill = gene, label = gene,
                  forward = orientation)) +
  geom_gene_arrow(arrowhead_height = unit(10, "mm"), arrowhead_width = unit(2, "mm"),
                  arrow_body_height = unit(10, "mm")) +
  geom_gene_label() +
  geom_feature(
    data = filtered.gaps,
    aes(x = position, y = molecule)
  ) +
  geom_feature_label(
    data = filtered.gaps,
    aes(x = position, y = molecule, label = name)
  ) +
  #facet_wrap(~ molecule, scales = "free", ncol = 1) +
  scale_fill_manual(values = myColors) +
  theme_genes() +
  scale_x_continuous(limits = c(0,26000), breaks = c(0,6500,13000,19500,26000))
plot2

```

```{r Nod locus - subset export, results='hide', message=FALSE, warning=FALSE}

# Save the plot 
pdf("nod.locus.subset.pdf", width=15, height=4)
plot2
dev.off()

# Clean the workspace
rm(list = ls())

```

# Symbiovar distribution

Lastly, summarize the distribution of symbiovar (subclades) across species

```{r Symbiovar distribution, message=FALSE, warning=FALSE}

# Load the data
data <- read.table("genome_stats.txt",
                    header = TRUE, row.names = NULL, sep="\t")

# Make the table
symbiovar.stats <- data %>%
  group_by(Species, Symbiovar) %>%
  summarize(Count = n())
symbiovar.stats.wide <- symbiovar.stats %>%
  pivot_wider(names_from = Symbiovar, values_from = Count) %>%
  replace(is.na(.), 0)

# Create a table rounded to 0 decimal place
kable(symbiovar.stats.wide, align = "l",
      caption = "Symbiovar distribution") %>%
  kable_classic(full_width = FALSE)

```