---
title: "Nitrogen_addition"
output: html_document
date: "2026-02-23"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(width=200)

```

# Screening a subset of isolates on navy bean

```{r Load libraries, results='hide', message=FALSE, warning=FALSE}

library(lme4)
library(phyloseq)
library(car)
library(emmeans)
library(ggplot2)
library(patchwork)
library(vegan)
library(dplyr)
library(tidyr)
library(kableExtra)
library(tibble)
library(rcartocolor)
library(multcomp)
library(tidyverse)

```

# Experiment set up

Navy bean plants were grown in a greenhouse with and without a rhizobium inoculant (strain QUR00071). Plants were grown with two types of nitrogen supplemented treatments: one time application at the beginning of the experiment, and continuous application of nitrogen (nitrogen applied once a week). Several different nitrogen concentrations were tested in the experiment. 

Plants were grown for 4 weeks. At the end of the experiment, shoot biomass was dried and weighed. 

```{r Load data, results='hide', message=FALSE, warning=FALSE}

# load the data 
data <- read.table("nitrogen_addition_data.txt",
                   header = TRUE, row.names = NULL)

# Remove one plant that sprouted a week and half late and other plant without SDW data
data <- data %>%
  filter(Sample!=28) %>%
  filter(Sample!=32)

```

# Statistical analysis

## Test the effect of one-time nitrogen addition

```{r Stats - once, message=FALSE, warning=FALSE}

# Filter the dataset for the one time data 
data.once<- data %>%
  filter(Nitrogen_mM=="0" | Treatment=="Once") %>%
  mutate(Rhizobia=as.factor(Rhizobia))

data.once$Nitrogen_of <- factor(data.once$Nitrogen_mM, levels = c(0,2.5,5,10,20),
                              ordered = TRUE)
data.once$Rhizobia <- factor(data.once$Rhizobia, levels = c("Yes", "No"))

# linear model 
model.once <- lm(log(Biomass_g) ~ Rhizobia * Nitrogen_of, data = data.once)
layout(matrix(1:4, ncol = 2))
plot(model.once)
summary(model.once)
Anova(model.once, type=3)

# Get contrasts
model.once.em <- emmeans(model.once, ~ Rhizobia | Nitrogen_of,
                         infer = TRUE, type = "response")
model.once.em.df <- as.data.frame(model.once.em)
model.once.em.rg <- regrid(model.once.em)
model.once.em.rg.con <- contrast(model.once.em.rg, by = "Nitrogen_of", 
                                 method = "pairwise", infer = TRUE,
                                 ratios = FALSE, type = "response")
model.once.em.rg.con.df <- as.data.frame(model.once.em.rg.con)

# Create a table rounded to 2 decimal place
kable(model.once.em.rg.con.df, align = "l", digits = 2,
      caption = "One-time nitrogen summary") %>%
  kable_classic(full_width = FALSE)

```

There is a significant effect of inoculation and the interaction of inoculation with nitrogen. Now plot the contrasts.

```{r Stats - once - plot contrasts, message=FALSE, warning=FALSE}

# Make plot
plot.con.once <- ggplot(model.once.em.rg.con.df, aes(x = Nitrogen_of, y = estimate)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  geom_line(aes(group = contrast)) +
  scale_x_discrete(name = "Nitrogen concentration (mM)") +
  scale_y_continuous(name = "SDW change with inoculation (g/plant)",
                     limits = c(0,3),
                     breaks = c(0,0.75,1.5,2.25,3)) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12), 
        legend.position = "right", 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))  
plot.con.once

```

## Test the effect of ongoing nitrogen addition

```{r Stats - ongoing, message=FALSE, warning=FALSE}

# filter the dataset 
data.ongoing<- data %>%
  filter(Nitrogen_mM=="0" | Treatment=="Ongoing") %>%
  mutate(Rhizobia=as.factor(Rhizobia))

data.ongoing$Nitrogen_of <- factor(data.ongoing$Nitrogen_mM, levels = c(0,1,2.5,10,20),
                              ordered = TRUE)
data.ongoing$Rhizobia <- factor(data.ongoing$Rhizobia, levels = c("Yes", "No"))

# linear model 
model.ongoing <- lm(log(Biomass_g) ~ Rhizobia * Nitrogen_of, data = data.ongoing)
plot(model.ongoing)
summary(model.ongoing)
Anova(model.ongoing, type=3)

# Get contrasts
model.ongoing.em <- emmeans(model.ongoing, ~ Rhizobia | Nitrogen_of,
                         infer = TRUE, type = "response")
model.ongoing.em.df <- as.data.frame(model.ongoing.em)
model.ongoing.em.rg <- regrid(model.ongoing.em)
model.ongoing.em.rg.con <- contrast(model.ongoing.em.rg, by = "Nitrogen_of", 
                                 method = "pairwise", infer = TRUE,
                                 ratios = FALSE, type = "response")
model.ongoing.em.rg.con.df <- as.data.frame(model.ongoing.em.rg.con)

# Create a table rounded to 2 decimal place
kable(model.ongoing.em.rg.con.df, align = "l", digits = 2,
      caption = "Ongoing nitrogen summary") %>%
  kable_classic(full_width = FALSE)

```

There is a significant effect of inoculation, nitrogen, and the interaction of inoculation with nitrogen. Now plot the contrasts.

```{r Stats - ongoing - plot contrasts, message=FALSE, warning=FALSE}

# Make plot
plot.con.ongoing <- ggplot(model.ongoing.em.rg.con.df, aes(x = Nitrogen_of, y = estimate)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  geom_line(aes(group = contrast)) +
  scale_x_discrete(name = "Nitrogen concentration (mM)") +
  scale_y_continuous(name = "SDW change with inoculation (g/plant)",
                     limits = c(-2.5,2.5),
                     breaks = c(-2.5,-1.25,0,1.25,2.5)) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12), 
        legend.position = "right", 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))  
plot.con.ongoing

```

# Data visualization

```{r Create plots - ongoing, message=FALSE, warning=FALSE}

# Get stats for the data
data.ongoing$Rhizobia <- factor(data.ongoing$Rhizobia, levels = c("No", "Yes"))
stats.ongoing <- data.ongoing %>%
  group_by(Nitrogen_mM, Rhizobia) %>%
  summarize(Mean = mean(Biomass_g),
            SD = sd(Biomass_g)) %>%
  mutate(Nitrogen=as.factor(Nitrogen_mM))
stats.ongoing

# Plot date for ongoing application
ongoing.plot <- ggplot(stats.ongoing, aes(x=Nitrogen, y=Mean, fill=Rhizobia)) + 
  scale_fill_manual(values=c("white", "lightgrey")) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           colour = "black") +
  geom_errorbar(aes(ymin = Mean-SD, ymax = Mean+SD),
                position = position_dodge(width = 0.9),
                stat = "identity", width = 0.5, linewidth = 0.5) +
  geom_vline(xintercept = c(1.5,2.5,3.5,4.5), color="lightgrey", linetype = "longdash") +
  scale_x_discrete(name = "Nitrogen concentration (mM)") +
  scale_y_continuous(name = "Shoot dry weight (SDW) (g/plant)",
                     limits = c(0,6),
                     breaks = c(0,1.5,3,4.5,6)) +
  theme_bw() + 
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12), 
        legend.position = "none", 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))  
ongoing.plot

```

```{r Create plots - one time, message=FALSE, warning=FALSE}

# Get stats for the data
data.once$Rhizobia <- factor(data.once$Rhizobia, levels = c("No", "Yes"))
stats.once <- data.once %>%
  group_by(Nitrogen_mM, Rhizobia) %>%
  summarize(Mean = mean(Biomass_g),
            SD = sd(Biomass_g)) %>%
  mutate(Nitrogen=as.factor(Nitrogen_mM))
stats.once

# Plot date for onetime application
once.plot <- ggplot(stats.once, aes(x=Nitrogen, y=Mean, fill=Rhizobia)) + 
  scale_fill_manual(values=c("white", "lightgrey")) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           colour = "black") +
  geom_errorbar(aes(ymin = Mean-SD, ymax = Mean+SD),
                position = position_dodge(width = 0.9),
                stat = "identity", width = 0.5, linewidth = 0.5) +
  geom_vline(xintercept = c(1.5,2.5,3.5,4.5), color="lightgrey", linetype = "longdash") +
  scale_x_discrete(name = "Nitrogen concentration (mM)") +
  scale_y_continuous(name = "Shoot dry weight (SDW) (g/plant)",
                     limits = c(0,6),
                     breaks = c(0,1.5,3,4.5,6)) +
  theme_bw() + 
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12), 
        legend.position = "none", 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))  
once.plot

```

```{r Export plots, results='hide', message=FALSE, warning=FALSE}

# Save the plot 
pdf("nitrogen.addition.pdf", width=12, height=8)
(ongoing.plot + plot.con.ongoing + plot_layout(widths = c(1.5, 1))) / 
  (once.plot + plot.con.once + plot_layout(widths = c(1.5, 1)))
dev.off()

```
